---
title: "Intro to lehdr: Grabbing Longitudinal and Employer-Household Dynamics (LEHD) data!"
author: "Jasmine Siyu Wu"
date: "2023-09-20"
output: 
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    code_download: true
    code_folding: "hide"
editor_options: 
  markdown: 
    wrap: 72
---


## Introduction


`lehdr` is a very simple yet important R package to city planners and urban spatial analysts. 



Begin your tutorial with an introduction to the R package. Provide some background information about the package, its purpose, and why it is useful.

Grab Longitudinal Employer-Household Dynamics (LEHD) Flat Files

Authors: Jamaal Green [cre, aut] , Liming Wang [aut] , Dillon Mahmoudi [aut] , Matthew Rogers [ctb] , Kyle Walker [ctb]

</br>

----------------------------------------------


## Installation

`lehdr` has not yet been submitted to CRAN, so we need to install the package using `devtools`. Additionally, we will also use other packages to achieve common goals of using the LODES data and to assist visualisations. Packages include `ggplot2`, `stringr`, `dplyr`, `kableExtra`, `tidycensus`, `leaflet`, `shiny`, `sf`, and `purrr`.

```{r install packages, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# install other packages used in this tutorial
install.packages(setdiff(c("ggplot2", "stringr", "dplyr", "devtools",
                           "kableExtra", "tidycensus", "leaflet", "leaflet.extras",
                           "leaflet.providers", "shiny", "sf", "purrr"),
                rownames(installed.packages())),
                repos="http://cran.rstudio.com")

library(dplyr)
library(stringr)
library(devtools)
library(kableExtra)
library(tidycensus)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
library(shiny)
library(sf)
library(purrr)
library(ggplot2)


# install and load the lehdr package!!
devtools::install_github("jamgreen/lehdr")
library(lehdr)

```

----------------------------------------------


## Usage & Demonstration

There are two external functions - `grab_lodes()` and `grab_crosswalk()` - and one internal function -`aggregate_lodes_df()` - that supports the user-end functions in the `lehdr` package. In this tutorial, we will focus on the `grad_lodes()` function since the `grab_crosswalk()` function often results in error messages^[I failed to run this function.].

The `grav_lodes` function is to download LODES origin-destination ("od"), residential association ("rac"), or workplace association ("wac") tables, which is identified by the `lodes_type` parameter. Using the `job_type` parameter, we can choose the job types that we want to grab from the LODES data: "JT00" for all jobs, "JT01" for Primary Jobs, "JT02" for All Private Jobs, "JT03" for Private Primary jobs, "JT04" for All Federal jobs, "JT05" for Federal Primary jobs. The default spatial unit of job aggregation (`agg_geo`) is Census block, and other options include "bg" for Census block group, "tract" for Census tract, "county" for county, or "state" for state. More parameters can be found in the package manual document^[lehdr PDF Manual: https://cran.utstat.utoronto.ca/web/packages/lehdr/lehdr.pdf]. 


### Use `grab_lodes` to grab LODES origin-destination tables

For example, we can crab the county-level origin-destination data for Pennsylvania in 2021. 
In this example, we are interested in only primary jobs and all types of segments of the workforce.

```{r grab lodes PA, warning = FALSE, message = FALSE, include=TRUE}

# download and load 2021 O-D data for Pennsylvania and aggregate
# to the county level, only downloading primary jobs
od_PA_county_2021 <- grab_lodes(
                                state='pa',
                                year=2019,
                                lodes_type = "od",
                                job_type = "JT01",
                                agg_geo = "county",
                               )

# show the first 10 rows of OD pairs in the above retrieved dataframe
head(od_PA_county_2021, n=10) %>% 
  rename("Year" = year,
         "State" = state,
         "Workplace county geoid" = w_county,
         "Residence county geoid" = h_county,
         "Total number of jobs" = S000, 
         "# jobs for workers <= 29" = SA01,
         "# jobs for workers 30 - 54" = SA02,
         "# jobs for workers >= 55" = SA03,
         "# jobs with monthly earnings <= $1250" = SE01, 
         "# jobs with monthly earnings $1251-3333" = SE02,
         "# jobs with monthly earnings >= $3333" = SE03,
         "# jobs in Goods Producing industry sectors" = SI01,
         "# jobs in Trade, Transportation, and Utilities industry sectors" = SI02,
         "# jobs in All Other Services industry sectors" = SI03 )%>% 
  kable(., "html", digits=1, align = "lccrr",
        caption = "") %>% 
  kable_styling() %>%
  scroll_box(width = "800px", height = "600px")
```

Now we have a dataframe of job origin-destination pairs of Pennsylvania counties. To know and visualize total number of jobs located in each county, we need to first aggregate jobs by `w_county`. The table below shows the top 10 counties with the most jobs.

```{r total jobs PA counties, warning = FALSE, message = FALSE, include=TRUE}

# aggregate jobs by w_county
# to get the total primary jobs at county level
totjobs_PA_county_2021 <- od_PA_county_2021 %>% 
  group_by(w_county) %>%
  summarise(totjobs = sum(S000))


# show first ten rows in the dataframe
head(totjobs_PA_county_2021, n=10) %>% 
  arrange(desc(totjobs)) %>%
  rename("Total primary jobs" = totjobs,
         "Workplace county geoid" = w_county) %>% 
  kable(., "html", digits=1, align = "lccrr", 
        format.args = list(big.mark = ","), caption = "Total 10 Pennsylvania counties with most primary jobs, 2021") %>% 
  kable_styling(position = "center")

```

To do this workplace job aggregation faster, we can use `grab_lodes()` directly to grab workplace association ("wac") tables. The wac tables also provide us with more detailed job categories, such as industry sectors defined in the NAICS system. The following code is an example of grabing workpace job association table for Pennsylvania counties in 2021, in which `C000` indicates the total number of jobs (primary and others). See more details about fields in the LODES manual^[LODES Public Data Structure: https://lehd.ces.census.gov/data/lodes/LODES7/LODESTechDoc7.5.pdf].

```{r grab lodes PA wac, warning = FALSE, message = FALSE, include=TRUE}

# download and load 2021 wac data for Pennsylvania and aggregate
# to the county level, only downloading primary jobs
wac_PA_county_2021 <- grab_lodes(
                                state='pa',
                                year=2019,
                                lodes_type = "wac",
                                job_type = "JT01",
                                agg_geo = "county",
                               )

# show the first 10 rows of wac data
head(wac_PA_county_2021, n=10) %>% 
  kable(., "html", digits=1, align = "lccrr",
        caption = "") %>% 
  kable_styling() %>%
  scroll_box(width = "800px", height = "600px")
```



### Integrating the LODES data with `sf` objects for visulization

(Authors'next steps: integration of the package with the `sf` and `tigris` packages to allow for easier mapping of LODES data)


Yet, we don't know what counties are and where these counties are from the LODES tables, in which no familiar place names and geometries are provided. We can add more richness to our LODES data using the `tidycensus` package to attach Census geographies' basic information. Then, `ggplot` is used to visualize the Pennsylvania counties and their total number of primary jobs in 2021.


```{r total jobs PA counties sf, warning = FALSE, message = FALSE, include=TRUE, results=FALSE}
# get counties and their geometries from 2021 Census data
PA_county_2021 <- get_acs(geography = "county", 
                          state= 42, #PA
                          year = 2021,
                          variables = c('B08006_001'),
                          survey = "acs5",
                          output = "wide",
                          geometry = TRUE)

totjobs_PA_county_2021_sf <- merge(x=PA_county_2021, y=totjobs_PA_county_2021, 
                                by.x="GEOID", by.y="w_county", all.x=TRUE)
```


Let's first visualize how many primary jobs are located in each county in Pennsylvania.

```{r total jobs PA counties map, warning = FALSE, message = FALSE, include=TRUE, fig.align='center', fig.width=10, fig.height=8}

#Map total primary jobs in Pennsylvania couties
pal <- leaflet::colorNumeric(viridis::viridis_pal(option = "F", direction = 1, 
                                                  begin = 1, end = 0)(5), 
                             domain = totjobs_PA_county_2021$totjobs)

totjobs_PA_county_2021_sf %>% 
  mutate(label = paste0("<b>", NAME, ":</b> ", round(totjobs))) %>%
  leaflet(.) %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    color = "transparent", weight = 0.1, opacity = 1,
    fillColor = ~pal(totjobs), fillOpacity = 0.7,
    label = ~lapply(label, HTML),
    labelOptions = labelOptions(direction = "top"),
    highlight = highlightOptions(
      color = "#FFF", bringToFront = TRUE
    )
  ) %>%
  addLegend(
    values = ~totjobs, opacity = 0.7,
    pal = pal,
    title = "Total primary jobs", 
    position = "topleft"
  )

```

### Visualize job flows using the LODES OD data

Another question that we can answer using the LODES data is people's commuting pattern.


```{r pa counties od map, warning = FALSE, message = FALSE, include=TRUE, fig.align='center', fig.width=10, fig.height=8}

# generate centroids of Philadelphia counties
centroids_PA_county_2021<- st_centroid(PA_county_2021) %>%
  st_transform(st_crs(PA_county_2021)) %>%
  mutate(lon = map_dbl(geometry, ~st_centroid(.x)[[2]]),
         lat = map_dbl(geometry, ~st_centroid(.x)[[1]]))


# join x y coordinates to the origin and then the destination points.
or_PA_county_2021 <- merge(od_PA_county_2021, centroids_PA_county_2021, by.x="h_county", by.y="GEOID") %>% 
  rename(or_lon = "lon",
         or_lat = "lat")
dest_PA_county_2021 <-  merge(or_PA_county_2021, centroids_PA_county_2021, by.x="w_county", by.y="GEOID")%>% 
  rename(dest_lon = "lon",
         dest_lat = "lat")


# plot 
ggplot(dest_PA_county_2021, aes(oX, oY))+
  #The next line tells ggplot that we wish to plot line segments. The "alpha=" is line transparency and used below
  geom_segment(aes(x=or_lat, y=or_lon, xend=dest_lat, yend=dest_lon, alpha=S000), col="white")+
  #Here is the magic bit that sets line transparency - essential to make the plot readable
  scale_alpha_continuous(range = c(0.03, 0.3))+
  #Set black background, ditch axes and fix aspect ratio
  theme(panel.background = element_rect(fill='black',colour='black'))+
  scale_x_continuous("", breaks=NULL) +
  scale_y_continuous("", breaks=NULL) +
  coord_equal()

```



----------------------------------------------

## Conclusion

Summarize the key points covered in your tutorial.
Encourage users to explore the package further and provide additional resources if available.







